---
title: "Model Diagnostics & Spatial Autocorrelation"
subtitle: "Week 7: CPLN 5920-MUSA 5080"
author: "Dr. Allison Lassiter"
date: "March 2, 2026"
format: 
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    code-line-numbers: true
    incremental: false
    smaller: true
    scrollable: true
---

# Homework Feedback & Tips

## Before We Start: A Quick Note on Lab 2 Submissions

Many of you have **messy output** in your rendered HTML files from `tigris` and `tidycensus` functions.

**Example of what we're seeing:**
```
Retrieving data for the year 2022
  |======================================================================| 100%
  |======================================================================| 100%
Downloading: 4.3 MB     
Downloading: 3.7 MB
```

**This clutters your professional report!**

---

## The Problem: Progress Bars in Rendered Output {.smaller}

**What's happening:**

When you use `tigris` or `tidycensus` functions, they show download progress by default.

:::: {.columns}
::: {.column width="50%"}
### **In your console (good!):**
```{r}
#| eval: false
#| echo: true
library(tidycensus)
tracts <- get_acs(
  geography = "tract",
  variables = "B01003_001",
  state = "PA",
  geometry = TRUE
)
```

Shows:
```
Getting data from the 2018-2022 5-year ACS
  |======| 100%
```

**This is helpful when coding!**
:::

::: {.column width="50%"}
### **In your rendered HTML (bad!):**

All those progress messages appear as ugly text in your final document.

**This looks unprofessional** and makes your work harder to read.

**Solution:** Suppress progress messages in your code chunks
:::
::::

---

## The Solution: Add `progress = FALSE` {.smaller}

**Two ways to fix this:**

### **Option 1: In each function call**
```{r}
#| eval: false
#| echo: true
# Add progress = FALSE to EVERY tigris/tidycensus call
tracts <- get_acs(
  geography = "tract",
  variables = "B01003_001",
  state = "PA",
  geometry = TRUE,
  progress = FALSE  # <-- Add this!
)

roads <- roads(state = "PA", 
               county = "Philadelphia",
               progress = FALSE)  # <-- Add this!
```

### **Option 2: Set globally at top of document** (Recommended!)
```{r}
#| eval: false
#| echo: true
# Add this near the top of your .qmd after loading libraries
options(tigris_use_cache = TRUE)
options(tigris_progress = FALSE)  # Suppress tigris progress bars
```

---

## Suggestion: Re-Render  {.smaller}


**Please go back to your homework and:**

1. Open your `.qmd` file
2. Add `progress = FALSE` to all `get_acs()`, `get_decennial()`, and `tigris` function calls
   - OR add the global options at the top of your document
3. **Re-render** your document (Click "Render" button)
4. Check that the HTML output is clean


---

# Today's Plan

## Agenda Overview

**Part 1: Review & Connect**

- Where we've been and where we're going
- The regression workflow so far

**Part 2: More Model Quality Evaluation**

- Train/test splits vs. cross-validation review
- Spatial patterns in errors
- Introduction to spatial autocorrelation

**Part 3: Moran's I as a Diagnostic**

- Understanding spatial clustering
- Calculating and interpreting Moran's I
- Local vs. global measures


**Part 4: Midterm Work Session**

---

# Part 1: Where We Are

## The Journey So Far

**Weeks 1-4:** Data foundations

- Census data, tidycensus, spatial data basics
- Visualization and exploratory analysis
- Data cleaning

**Week 5:** Linear regression fundamentals

- Y = f(X) + ε framework
- Train/test splits, cross-validation
- Checking assumptions

**Week 6:** Expanding the toolkit

- Categorical variables, polynomials, and interactions
- Spatial features (buffers, kNN, distance)
- Neighborhood fixed effects

## Last Week's Key Innovation

**You learned to create spatial features:**

```{r}
#| eval: false
#| echo: true

# Buffer aggregation
crimes_500ft <- count_features_in_buffer(houses, crimes, 500)

# k-Nearest Neighbors
crime_nn3 <- average_distance_to_knn(houses, crimes, k=3)

# Dummy variable and Fixed effects
lm(SalePrice ~ ... + as.factor(neighborhood))
```

**Today's Question:**

> *How do we know if our model has spatial structure in its errors?*

If errors are spatially clustered, we're missing something important!

## The Regression Workflow (Updated)

::: {.columns}
::: {.column width="50%"}
**Building the model:**

1. Visualize relationships
2. Engineer features
3. Fit the model
4. Evaluate performance (RMSE, R²)
5. Check assumptions
:::

::: {.column width="50%"}
**NEW: Spatial diagnostics:**

6. **Are errors random or clustered?**
7. **Do we predict better in some areas?**
8. **Is there remaining spatial structure?**
:::
:::

::: {.callout-important}
## Why This Matters

If errors cluster spatially, it suggests:

- Missing spatial variables
- Misspecified relationships
- Non-stationarity (relationships vary across space)
:::

---

# Part 2: Understanding Spatial Patterns in Errors

## What Are Model Errors?

For $$Y_i = f(X_i) + \epsilon_i$$


<br><br>

**Prediction error** for observation *i*:

$$e_i = \hat{y}_i - y_i$$

Where:

- $\hat{y}_i$ = predicted value
- $y_i$ = actual value

---

**In our house price context:**

```{r}
#| eval: true
#| echo: true

# Load packages and data
library(sf)
library(here)
library(tidyverse)

# Load Boston housing data
boston <- read_csv(here("data/boston.csv"))

# Simple model: Predict price from living area
baseline_model <- lm(SalePrice ~ LivingArea, data = boston)
summary(baseline_model)


boston_test <- boston %>%
  mutate(
    predicted = predict(baseline_model, boston),
    error = predicted - SalePrice,
    abs_error = abs(error),
    pct_error = abs(error) / SalePrice
  )

```

## Good Errors vs. Bad Errors

::: {.columns}
::: {.column width="50%"}
** Random errors (good)**

- No systematic pattern
- Scattered across space
- Prediction equally good everywhere
- Model captures key relationships
:::

::: {.column width="50%"}
** Clustered errors (bad)**

- Spatial pattern visible
- Under/over-predict in areas
- Model misses something about location
- Need more spatial features!
:::
:::

**How do we test this?**

Look for **spatial autocorrelation** in the errors

## Tobler's First Law (Revisited)

::: {.callout-note}
## The First Law of Geography

*"Everything is related to everything else, but near things are more related than distant things."*

— Waldo Tobler (1970)
:::

**Applied to house prices:**

- Nearby houses have similar prices
- Nearby neighborhoods have similar characteristics
- Crime in one block affects adjacent blocks

**Applied to model errors:**

- If nearby houses have similar, spatial errors...
- ...our model is missing a spatial pattern!
- Need to add more spatial features or fixed effects

## Visualizing Error Patterns

**Map your errors to see patterns:**

```{r}
library(sf)
options(scipen = 999)
# Convert boston data to sf object
boston_test <- boston_test %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform('ESRI:102286')  # MA State Plane (feet)


# Install if needed: install.packages("hexbin")
library(hexbin)

ggplot(boston_test) +
  geom_sf(aes(fill = error),
          shape = 21,
          size = 1,
          alpha = 0.6,
          stroke = 0.2) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-300000, 300000),
    oob = scales::squish
  ) +
  theme_void()
```

**What to look for:**

- Blue clusters (we under-predict)
- Red clusters (we over-predict)
- Random scatter (good!)

## Scatter Plot: Spatial Lag of Errors

**Create the spatial lag:**

```{r}
#| eval: true
#| echo: true
library(spdep)

# Define neighbors (5 nearest)
coords <- st_coordinates(boston_test)
neighbors <- knn2nb(knearneigh(coords, k=5))
weights <- nb2listw(neighbors, style="W") #"W" style gives neighbor average, most useful to us

# Calculate spatial lag of errors
boston_test$error_lag <- lag.listw(weights, boston_test$error)
```

**Then plot:**

```{r}
ggplot(boston_test, aes(x=error_lag, y=error)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="lm", color="red") +
  labs(title="Is my error correlated with neighbors' errors?",
       x="Avg error of 5 nearest neighbors",
       y="My error")
```

---

# Part 3: Moran's I

## What is Moran's I?

**Moran's I** measures spatial autocorrelation

> _"When I'm above/below average, are my neighbors also above/below average?"_

**Range:** -1 to +1

- **+1** = Perfect positive correlation (clustering)
- **0** = Random spatial pattern
- **-1** = Perfect negative correlation (dispersion)

**Formula (look's scary, but its so intuitive!):**

$$I = \frac{n \sum_i \sum_j w_{ij}(x_i - \bar{x})(x_j - \bar{x})}{\sum_i \sum_j w_{ij} \sum_i (x_i - \bar{x})^2}$$

Where $w_{ij}$ = spatial weight between locations *i* and *j*, <br>
$n$ is the number of observations 

## Worked Example: Understanding the Formula

**5 houses, predicting sale prices:**

| House | Actual Price | Predicted Price | Error | 
|-------|-------------|-----------------|-------|
| A     | $500k       | $400k          | +$100k |
| B     | $480k       | $400k          | +$80k  |
| C     | $420k       | $400k          | +$20k  |
| D     | $350k       | $400k          | -$50k  |
| E     | $330k       | $400k          | -$70k  |

**Mean error = +$16k**

**The question:** Are errors for nearby houses similar to each other?

---

## Step 1: Calculate Deviations from Mean

**Subtract the mean error from each house's error:**

| House | Error | Mean Error | Deviation from Mean |
|-------|-------|------------|-------------------|
| A     | +$100k | +$16k     | +$84k            |
| B     | +$80k  | +$16k     | +$64k            |
| C     | +$20k  | +$16k     | +$4k             |
| D     | -$50k  | +$16k     | -$66k            |
| E     | -$70k  | +$16k     | -$86k            |

**Positive deviation** = we over-predicted (actual > predicted)  
**Negative deviation** = we under-predicted (actual < predicted)

---

## Step 2: Multiply Neighbor Deviations

**For each neighbor pair, multiply their deviations:**

:::: {.columns}
::: {.column width="50%"}
**Neighbor Pairs:**

- A-B: $(+84k) \times (+64k) = +5,376$
- B-C: $(+64k) \times (+4k) = +256$
- C-D: $(+4k) \times (-66k) = -264$
- D-E: $(-66k) \times (-86k) = +5,676$

**Sum of products = 11,044**
:::

::: {.column width="50%"}
**What does this mean?**

**Positive products** = similar neighbors
- A-B: both over-predicted (both positive)
- D-E: both under-predicted (both negative)

**Negative product** = dissimilar neighbors  
- C-D: one over, one under

:::
::::

---


**Result:**

- Lots of positive products → **High Moran's I** (clustering)
- Products near zero → **Low Moran's I** (random)
- Negative products → **Negative Moran's I** (rare with errors)

---


## Computing Moran's I

**Calculate Moran's I for your errors:**

```{r}
#| eval: true
#| echo: true
# Test for spatial autocorrelation in errors
moran_test <- moran.mc(
  boston_test$error,        # Your errors
  weights,                  # Spatial weights matrix
  nsim = 999                # Number of permutations
)

# View results
moran_test$statistic         # Moran's I value
moran_test$p.value          # Is it significant?
```

**Interpretation:**

- **I > 0** and *p* < 0.05 → Significant clustering
- **I ≈ 0** → Random pattern (good!)
- **I < 0** → Dispersion (rare with errors)

## Visualizing Significance

**Compare observed I to random permutations:**

```{r}
#| eval: true
#| echo: false
library(ggplot2)

moran_df <- data.frame(sim_I = moran_test$res[1:999])

ggplot(moran_df, aes(x = sim_I)) +
  geom_histogram(binwidth = 0.01, fill = "gray70", color = "white") +
  geom_vline(aes(xintercept = moran_test$statistic), 
             color = "#FA7800", linewidth = 1.5) +
  labs(
    title = "Observed vs. Expected Moran's I",
    subtitle = "Orange line = observed | Gray = random permutations",
    x = "Moran's I",
    y = "Count"
  ) +
  theme_minimal()
```

## What Moran's I Tells You

::: {.callout-important}
## Decision Framework

**If Moran's I is high (errors clustered):**

1. Add more spatial features
   - Try different buffer sizes
   - Include more amenities/disamenities
   - Create neighborhood-specific variables

2. Try spatial fixed effects
   - Neighborhood dummies
   - Grid cell dummies

3. Consider spatial regression models
   - Spatial lag model
   - Spatial error model
   - (Advanced topic, not covered today)
:::

**If Moran's I ≈ 0 (random errors):**

✅ Your model adequately captures spatial relationships!

---

## Defining "Neighbors"

**Different ways to define spatial relationships:**

::: {.columns}
::: {.column width="33%"}
**Contiguity**

- Polygons that share a border
- Queen vs. Rook
:::

::: {.column width="33%"}
**Distance**

- All within X meters
- Fixed threshold
:::

::: {.column width="33%"}
**k-Nearest**

- Closest k points
- Adaptive distance
:::
:::

**For point data (houses), use k-nearest neighbors**

```{r}
#| eval: true
#| echo: true
# Create 5-nearest neighbors
coords <- st_coordinates(boston_test)
nb <- knn2nb(knearneigh(coords, k=5))
weights <- nb2listw(nb, style="W")
```

## New term: Spatial Lag

**Spatial lag** = average value of neighbors 

::: {.callout-tip}
## Example: 5 houses

| House | Sale Price | 2 Nearest | Spatial Lag |
|-------|-----------|-----------|-------------|
| A     | $200k     | B, C      | $275k       |
| B     | $250k     | A, C      | $250k       |
| C     | $300k     | B, D      | $275k       |
| D     | $350k     | C, E      | $350k       |
| E     | $400k     | D         | $350k       |
:::

**In R:**

```{r}
#| eval: true
#| echo: true
boston$price_lag <- lag.listw(weights, boston$SalePrice)
```

## "What About Spatial Lag/Error Models?" {.smaller}

*"In my spatial statistics class, I learned about spatial lag and spatial error models for dealing with spatial autocorrelation. Why aren't we using those here?"*

:::: {.columns}
::: {.column width="50%"}
### **Spatial Econometrics Models**
**(Spatial Statistics Class)**

**Spatial Lag Model:** $Y_i = \rho WY + \beta X_i + \varepsilon$

**Spatial Error Model:** $Y_i = \beta X_i + \lambda W\varepsilon + \xi$

**Purpose:** 

- Causal inference with spatial spillovers
- Understanding neighbor effects
- Correct standard errors for hypothesis testing
- Cross-sectional analysis

**When to use:** Academic research on spillover effects, peer influence, regional economics
:::

::: {.column width="50%"}
### **Predictive Spatial Features**
**(This Class)**

**Our Approach:** 
$$Y_i = \beta_0 + \beta_1 X_i + \beta_2(\text{crimes}_{500ft}) + \beta_3(\text{dist}_{downtown}) + \varepsilon_i$$

**Purpose:**

- Out-of-sample prediction
- Forecasting new observations  
- Applied machine learning
- Generalization to new areas

**When to use:** Real estate prediction, housing market forecasting, policy planning
:::
::::

---

## Why Not Spatial Lag Models for Prediction? {.smaller}

**1. Simultaneity Problem**

- Including $WY$ (neighbor prices) creates circular logic
- My price affects neighbors → neighbors affect me

**2. Prediction Paradox**

- Need neighbors' prices to predict my price
- But for new developments or future periods, those prices **don't exist yet**
- Can't generalize to truly new areas

**3. Data Leakage in CV**

- Geographic CV holds out spatial regions
- Spatial lag "leaks" information from test set
- Artificially good performance that won't hold


---


### **Our Solution: Spatial Features of X (Not Y)**

Instead of modeling dependence in **Y** (prices), model proximity in **X** (predictors)

| ❌ Spatial Lag | ✅ Our Approach |
|----------------|-----------------|
| "Near expensive houses" | "Near low crime areas" |
| Uses neighbor **prices** | Uses neighbor **characteristics** |
| Circular logic | Causal mechanism |
| Can't predict new areas | Generalizes well |

**If Moran's I shows clustered errors:**

✅ **Add more spatial features** (different buffers, more amenities)  
✅ **Try neighborhood fixed effects**  
✅ **Use spatial cross-validation**

❌ Don't add spatial lag of Y for prediction purposes

::: {.callout-tip}
## **The Bottom Line**
Both approaches are valid for different goals! Match method to purpose: **inference** → spatial lag/error models; **prediction** → spatial features.
:::



---

# Summary & Next Steps

## Key Takeaways

**Spatial autocorrelation in errors indicates model misspecification**

**Moran's I is a diagnostic tool:**

- Global I: overall clustering
- Maps of residuals give clues to what you might be missing

**Iterative improvement:**

- Diagnose → Engineer features → Re-test → Repeat
- Document what you try!



## Resources

**Spatial autocorrelation:**
- [https://mgimond.github.io/Spatial/spatial-autocorrelation.html](https://mgimond.github.io/Spatial/spatial-autocorrelation.html)

**spdep package:**
- [https://r-spatial.github.io/spdep/](https://r-spatial.github.io/spdep/)

---


