---
title: "Lab 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Yusuf Khalil"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data:
- Pennsylvania county boundaries
- Pennsylvania hospitals (from lecture data)
- Pennsylvania census tracts

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'


library(sf)
library(tidyverse)
library(tigris)
library(tidycensus)
library(scales)
library(patchwork)
library(here)


# Load spatial data
pa_counties <- st_read(here("data/Pennsylvania_County_Boundaries.shp"))
districts <- st_read(here("data/districts.geojson"))
hospitals <- st_read(here("data/hospitals.geojson"))
census_tracts <- tracts(state = "PA", cb = TRUE)
metro_areas <- core_based_statistical_areas(cb = TRUE)


# Check that all data loaded correctly

st_crs(hospitals)
head(hospitals)
count(hospitals)
distinct(hospitals)


count(census_tracts)
st_crs(census_tracts)

st_crs(pa_counties)

```

**Questions to answer:**
- How many hospitals are in your dataset? 222
- How many census tracts? 3445
- What coordinate reference system is each dataset in? Hospitals: WGS 84, Census Tracts: NAD83, Counties:  WGS 84 / Pseudo-Mercator 

---

#### Step 2: Get Demographic Data 

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:**
- Total population
- Median household income
- Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'


# Get demographic data from ACS
acs_vars_2023 <- load_variables(2023, "acs5", cache = TRUE)

old_people <- c("B01001A_014", "B01001A_015", "B01001A_016", "B01001A_029", "B01001A_030", "B01001A_031",
                "B01001B_014", "B01001B_015", "B01001B_016", "B01001B_029", "B01001B_030", "B01001B_031",
                "B01001D_014", "B01001D_015", "B01001D_016", "B01001D_029", "B01001D_030", "B01001D_031")

demp_data <- get_acs(geography = "Tract",
                     variables = c(Total_population ="B01003_001",
                                   Median_household_income = "B19013_001"
                                   ),
                     state = "PA",
                     year = 2023,
                     survey = "acs5",
                     output = "wide"
                     )

demo_data1 <- get_acs(geography = "Tract",
                     variables = old_people,
                     state = "PA",
                     year = 2023,
                     survey = "acs5",
                                                  
                     )%>%
  group_by(GEOID, NAME) %>%
  summarise(pop_65up = sum(estimate))%>%
  ungroup()
#Does not work when output is wide

demo_data_total <- inner_join(demp_data, demo_data1)
view(demo_data_total)

# Join to tract boundaries

demo_data_final <- census_tracts %>%
  inner_join(demo_data_total, by = "GEOID")

missing_income <- demo_data_final %>%
  filter(is.na(Median_household_incomeE))
missing_income

median_income <- demo_data_final %>%
  summarise(median_income = median(Median_household_incomeE, na.rm = TRUE))



```

**Questions to answer:**
- What year of ACS data are you using? 2023
- How many tracts have missing income data? 65
- What is the median income across all PA census tracts? $72,943.5

---

#### Step 3: Define Vulnerable Populations 

Identify census tracts with vulnerable populations based on TWO criteria:
1. Low median household income (choose an appropriate threshold)
2. Significant elderly population (choose an appropriate threshold)

**Your Task:**
```{r}
# Filter for vulnerable tracts based on your criteria

hist(demo_data_final$Median_household_incomeE)
quantile(demo_data_final$Median_household_incomeE, na.rm = TRUE)

hist(demo_data_final$pop_65up)
quantile(demo_data_final$pop_65up)

vulnerable_tracts <- demo_data_final %>%
  mutate(
    vuln_income = case_when(
      Median_household_incomeE <= 57863 ~ "vuln_income"),
    vuln_age = case_when(
      pop_65up >= 918 ~ "vuln_age"
    )
    )

total_vuln_tracts <- vulnerable_tracts %>% 
  filter(vuln_income == vuln_income | vuln_age == vuln_age) 


```

**Questions to answer:**
- What income threshold did you choose and why? $57,863 — the 25th percentile of median household income across PA tracts, identified using the histogram and quantile function. Tracts below this threshold represent the lowest-earning quarter of the state.

- What elderly population threshold did you choose and why?  918 — the 75th percentile of population aged 65+, meaning tracts above this have a disproportionately large elderly population compared to most PA tracts.

- How many tracts meet your vulnerability criteria? 1,625
- What percentage of PA census tracts are considered vulnerable by your definition? 47%

---

#### Step 4: Calculate Distance to Hospitals 

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Transform to appropriate projected CRS
vulnerable_tracts_proj <- vulnerable_tracts %>%
  filter(vuln_income == "vuln_income" | vuln_age == "vuln_age") %>%
  st_transform(crs = 2272)

hospitals_proj <- hospitals %>%
  st_transform(crs = 2272)

# Calculate distance from each tract centroid to nearest hospital

#Turn Vulnerable Tract centroids into points
vuln_tract_centroids <- vulnerable_tracts_proj %>%
  st_centroid()

#Finding distance to hosptial matrix
dist_to_hosp <- st_distance(vuln_tract_centroids, hospitals_proj) 
  apply(dist_to_hosp, 1, min)

dist_to_hosp

#Changing from feet to miles and rouding to 2 D.P
vuln_tract_centroids$nearest_hospital_miles <- round(apply(dist_to_hosp, 1, min) / 5280 ,2)

avg_dist_hosp <- mean(vuln_tract_centroids$nearest_hospital_miles)
avg_dist_hosp
view(vuln_tract_centroids)


```

**Requirements:**
- Use an appropriate projected coordinate system for Pennsylvania
- Calculate distances in miles
- Explain why you chose your projection

**Questions to answer:**
- What is the average distance to the nearest hospital for vulnerable tracts? 3.93
- What is the maximum distance? 19.16
- How many vulnerable tracts are more than 15 miles from the nearest hospital? 39

---

#### Step 5: Identify Underserved Areas 

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**
```{r}
# Create underserved variable
underserved_tract <- vuln_tract_centroids %>%
  mutate(
    underserved = case_when(
      nearest_hospital_miles > 15 ~ "Underserved"
    )
  )

underserved_tract %>% filter(underserved == "Underserved") %>% nrow()

```

**Questions to answer:**
- How many tracts are underserved? 39
- What percentage of vulnerable tracts are underserved? 2.4%
- Does this surprise you? Why or why not? No. Pennsylvania is a large state with both dense urban areas and sparsely populated rural regions. Most hospitals are concentrated in cities and suburban centers, so it's expected that a small number of rural tracts with vulnerable populations fall outside a 15-mile service radius.

---

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**
```{r}
# Spatial join tracts to counties
pa_counties_proj <- pa_counties %>% st_transform(crs = 2272)

underserved_tract_join <- st_join(underserved_tract, pa_counties_proj)

# Aggregate statistics by county
county_stats <- underserved_tract_join %>%
  st_drop_geometry(.) %>%
  group_by(COUNTY_NAM) %>%
  summarise(
    vulnerable_tracts = n(),
    underserved_tracts = sum(underserved == "Underserved", na.rm = TRUE)
  )

#County-level Statistics Calculations
county_stats_summary <- sum(county_stats$vulnerable_tracts)
avg_dist_nrst_hosp <- mean(underserved_tract$nearest_hospital_miles)
total_vuln_pop <- sum(underserved_tract$Total_populationE)
total_underserved <- sum(county_stats$underserved_tracts)

underserved_tract %>%
  filter(underserved == "Underserved") %>%
  group_by(NAMELSADCO) %>%
  summarise(total_vuln_pop = sum(Total_populationE, na.rm = TRUE)) %>%
  arrange(desc(total_vuln_pop))


#Visualisation for sanity

hospitals_final <- st_join(hospitals_proj, vulnerable_tracts_proj, left = FALSE)

underserved_tract_only <- vulnerable_tracts_proj %>%
  filter(GEOID %in% underserved_tract$GEOID[underserved_tract$underserved == "Underserved"])

hospital_buffer <- hospitals_final %>% 
  st_buffer(dist = 79200)

ggplot() +
  geom_sf(data = pa_counties, fill = "white", color = "gray") +
  geom_sf(data = hospital_buffer, fill = "lightblue", alpha = 0.4) +
  geom_sf(data = hospitals, color = "red", size = 2) +
  geom_sf(data = underserved_tract_only, fill = "pink", colour = "gray" ) +
  theme_void()

```

**Required county-level statistics:**
- Number of vulnerable tracts: 1625
- Number of underserved tracts: 39
- Percentage of vulnerable tracts that are underserved: 4/85
- Average distance to nearest hospital for vulnerable tracts: 3.96
- Total vulnerable population: 434055

**Questions to answer:**
- Which 5 counties have the highest percentage of underserved vulnerable tracts? Cameron County, Forest County, Sullivan County, Pike County, Potter County
- Which counties have the most vulnerable people living far from hospitals? Clearfield County, Snyder County, Chester County
- Are there any patterns in where underserved counties are located? Underserved counties are predominantly located in rural, northern and central Pennsylvania, areas with low population density, limited infrastructure, and greater distances between towns

---

#### Step 7: Create Summary Table 

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'asis'

library(kableExtra)
# Create and format priority counties table
underserved_tract_descend <- underserved_tract %>%
  arrange(desc(nearest_hospital_miles))

library(kableExtra)

underserved_tract_descend <- underserved_tract %>%
  st_drop_geometry() %>%
  arrange(desc(nearest_hospital_miles)) %>%
  head(10) %>%
  select(GEOID, NAMELSADCO, NAMELSAD, Total_populationE ,nearest_hospital_miles, underserved)

underserved_tract_descend %>%
  kable(
    format = "html",
    col.names = c("Tract GEOID", "County","Tract", "Total Population","Distance (mi)", "Status"),
    caption = "Top 10 Priority Counties for Healthcare Investment",
    digits = 2,
    align = "llcl"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Source: ACS 2023. Underserved = nearest hospital > 15 miles.")


```

**Requirements:**
- Use `knitr::kable()` or similar for formatting
- Include descriptive column names
- Format numbers appropriately (commas for population, percentages, etc.)
- Add an informative caption
- Sort by priority (you decide the metric)

---

## Part 2: Comprehensive Visualization 

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth 

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**
```{r}
# Create county-level access map
#Joining county level CT data to vulnerable tract data
underserved_tract <- st_drop_geometry(underserved_tract)
perc_vuln_trat <- left_join(demo_data_final, underserved_tract)

#Summarising number of CT per County
#Add filter so we only count number of VULNERABLE tracts
CT_per_County <- perc_vuln_trat %>% 
  filter(vuln_income == "vuln_income" | vuln_age == "vuln_age") %>% 
  group_by(NAMELSADCO) %>% 
  summarise(
    Number_of_ct = n()
  )

CT_per_County

vuln_tract_per_county <- perc_vuln_trat %>% 
  filter(underserved == "Underserved") %>% 
  group_by(NAMELSADCO) %>% 
  summarise(
    number_of_ct = n()
  )

vuln_tract_per_county

#Caclualting percentage of vulnerable tracts underserved 
CT_per_County <- st_drop_geometry(CT_per_County)
vuln_tract_per_county <- st_drop_geometry(vuln_tract_per_county)

perc_vuln_trat <- vuln_tract_per_county %>% 
  left_join(CT_per_County, by = "NAMELSADCO") %>% 
  mutate(perc_counties_underserved = round((number_of_ct / Number_of_ct)*100, 2))
perc_vuln_trat


#Gives pa counties with desirable geometry the correct county name so we can merge
county_choropleth <- pa_counties %>%
  mutate(NAMELSADCO = str_to_title(COUNTY_NAM),
         NAMELSADCO = paste(NAMELSADCO, "County")) %>%
  left_join(perc_vuln_trat, by = "NAMELSADCO")

 
#Mapping perc_vuln_tract somehow?

ggplot() +
  geom_sf(data = pa_counties, fill = "white", color = "gray") +
  geom_sf(data = county_choropleth, aes(fill = perc_counties_underserved), colour = "white") +
    scale_fill_viridis_c(
      name = "% Vulnerable Tract",
      option = "magma" ) +
  geom_sf(data = hospitals, color = "red", size = 1) +

  labs(
    title = "Percentage of Vulnerable Tracts Underserved by County",
    subtitle = "Red points = Hospital locations",
    caption = "Source: ACS 2023, PA Hospital Data. Underserved = >15 miles from nearest hospital."
  ) +
  theme_void()



```

**Requirements:**
- Fill counties by percentage of vulnerable tracts that are underserved: yes
- Include hospital locations as points: yes
- Use an appropriate color scheme: 
- Include clear title, subtitle, and caption
- Use `theme_void()` or similar clean theme
- Add a legend with formatted labels

---

### Map 2: Detailed Vulnerability Map 

Create a map highlighting underserved vulnerable tracts.

**Your Task:**
```{r}
# Create detailed tract-level map

ggplot() +
  geom_sf(data = pa_counties, fill = "white", color = "gray") +
  geom_sf(data = hospitals, color = "red", size = 2) +
  geom_sf(data = underserved_tract_only, fill = "pink", colour = "black", linewidth = 0.5 ) +
  labs(
    title = "Underserved Vulnerable Census Tracts in Pennsylvania",
    subtitle = "Red points = Hospitals, Pink tracts = Vulnerable and >15 miles from nearest hospital",
    caption = "Source: ACS 2023, PA Hospital Data"
  ) +
  theme_void()



```

**Requirements:**
- Show underserved vulnerable tracts in a contrasting color: Yes
- Include county boundaries for context: yes
- Show hospital locations: yes
- Use appropriate visual hierarchy (what should stand out?): yes
- Include informative title and subtitle: tbd

---

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**
```{r}
# Create distribution visualization

scatter_plot <- underserved_tract %>% 
  filter(vuln_income == "vuln_income" | vuln_age == "vuln_age")

ggplot(scatter_plot, aes(x = nearest_hospital_miles, y = Total_populationE)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red")+
  labs(x = "Distance", y = "Vulnerable Population Size") +
  theme_minimal()



```

-The scatter plot shows a weak positive relationship between distance to the nearest hospital and tract population size, with most vulnerable tracts clustered within 5 miles of a hospital. The nearly flat trend line suggests that distance to hospitals does not vary significantly with population size, though the sparse scatter beyond 15 miles highlights a small number of isolated tracts that may warrant targeted intervention.

---

## Part 3: Bring Your Own Data Analysis 

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Challenge Options

Choose ONE of the following challenge exercises, or propose your own research question using OpenDataPhilly data (https://opendataphilly.org/datasets/). 


---

### Your Analysis

**Your Task:**

1. **Find and load additional data**
   - Document your data source
   - Check and standardize the CRS
   - Provide basic summary statistics

```{r}
#| message: false
#| warning: false
#| results: 'hide'

library(sf)
library(tidyverse)
library(tigris)
library(tidycensus)
library(scales)
library(patchwork)
library(here)



#Load datasets

transit_stops <- st_read(here("data/Transit_Stops_(Spring_2025).shp"))
st_crs(transit_stops)
census_tracts2 <- tracts(state = "PA", cb = TRUE)

rail_lines <- c("MFL", "BSL", "NHSL")
trolley_lines <- c("10", "11", "13", "15", "34", "36", "101", "102")

septa_stops <- transit_stops %>%
  mutate(mode = case_when(
    LineAbbr %in% rail_lines ~ "Rail",
    LineAbbr %in% trolley_lines ~ "Trolley",
    TRUE ~ "Bus"
  ))

philly_tracts <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = c(
    total_pop = "B01003_001",
    white = "B02001_002",
    black = "B02001_003",
    asian = "B02001_005",
    hispanic = "B03003_003"
  ),
  year = 2022,
  geometry = TRUE,
  output = "wide"
)

septa_stops <- st_transform(septa_stops, st_crs(philly_tracts))

#summary statistics

stops_by_tract <- st_join(septa_stops, philly_tracts)

stops_count <- stops_by_tract %>%
  st_drop_geometry() %>%
  group_by(NAME) %>%
  summarise(num_stops = n())
stops_count


```

**Questions to answer:**
- What dataset did you choose and why? SEPTA transit stop locations across Philadelphia. Public transit access is a key determinant of economic opportunity and quality of life, particularly for low-income and minority communities who are more likely to depend on public transport.

- What is the data source and date? OpenDataPhilly (opendataphilly.org), SEPTA Routes & Stops Locations dataset, Spring 2025.
- How many features does it contain? 22,478
- What CRS is it in? Did you need to transform it? WGS 84 / Pseudo-Mercator,  I transformed it to match the ACS tract data using st_transform()

---

2. **Pose a research question**

How equitably are SEPTA transit stops distributed across Philadelphia's census tracts, and do disparities in access vary by racial composition?


---

3. **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):**
- Buffers
- Spatial joins
- Spatial filtering with predicates
- Distance calculations
- Intersections or unions
- Point-in-polygon aggregation

**Your Task:**
```{r}
# Your spatial analysis

#Distance from each tract centroid to nearest transit stop
philly_tracts_proj <- philly_tracts %>%
  st_transform(crs = 2272)

septa_stops_proj <- septa_stops %>%
  st_transform(crs = 2272)

#stops by census tract
stops_by_tract <- st_join(septa_stops, philly_tracts)

stops_count <- stops_by_tract %>%
  st_drop_geometry() %>%
  group_by(NAME) %>%
  summarise(num_stops = n())
stops_count

#join back to main data
philly_tracts_proj <- philly_tracts_proj %>%
  left_join(st_drop_geometry(stops_count), by = "NAME") %>%
  mutate(num_stops = replace_na(num_stops, 0),
         stops_per_1000 = round((num_stops / total_popE) * 1000, 2))

#make centroids
tract_centroids <- philly_tracts_proj %>%
  st_centroid()

dist_to_stop <- st_distance(tract_centroids, septa_stops_proj)

tract_centroids$nearest_stop_miles <- round(apply(dist_to_stop, 1, min) / 5280, 2)


#classify tracts by race
philly_race_analysis <- tract_centroids %>%
  mutate(
    majority_race = case_when(
      whiteE / total_popE > 0.5 ~ "Majority White",
      blackE / total_popE > 0.5 ~ "Majority Black",
      asianE / total_popE > 0.5 ~ "Majority Asian",
      hispanicE / total_popE > 0.5 ~ "Majority Hispanic",
      TRUE ~ "No Majority"
    )
  )

#summary statistics for race
race_summary <- philly_race_analysis %>%
  st_drop_geometry() %>%
  group_by(majority_race) %>%
  summarise(
    num_tracts = n(),
    avg_stops = round(mean(num_stops, na.rm = TRUE), 2),
    avg_stops_per_1000 = round(mean(stops_per_1000, na.rm = TRUE), 2),
    avg_dist_nearest_stop = round(mean(nearest_stop_miles, na.rm = TRUE), 2)
  )

race_summary

#map 1
ggplot() +
  geom_sf(data = philly_tracts_proj, aes(fill = num_stops), colour = "white", linewidth = 0.1) +
  scale_fill_viridis_c(name = "Number of Stops per Census Tract", option = "plasma") +
  labs(
    title = "SEPTA Transit Stop Density by Census Tract",
    subtitle = "Philadelphia County"
  ) +
  theme_void()

#map2
philly_race_map <- philly_tracts_proj %>%
  mutate(
    majority_race = case_when(
      whiteE / total_popE > 0.5 ~ "Majority White",
      blackE / total_popE > 0.5 ~ "Majority Black",
      asianE / total_popE > 0.5 ~ "Majority Asian",
      hispanicE / total_popE > 0.5 ~ "Majority Hispanic",
      TRUE ~ "No Majority"
    )
  )

septa_philly <- septa_stops_proj[philly_tracts_proj, ]

ggplot() +
  geom_sf(data = philly_race_map, aes(fill = majority_race), colour = "white", linewidth = 0.1) +
  geom_sf(data = septa_philly, size = 0.1, alpha = 0.3) +
  labs(
    title = "SEPTA Stop Locations by Racial Composition",
    subtitle = "Points = Transit Stops",
    fill = "Majority Race"
  ) +
  theme_void()

#Chart 1 - distance by race
ggplot(st_drop_geometry(philly_race_analysis), aes(x = majority_race, y = nearest_stop_miles, fill = majority_race)) +
  geom_boxplot() +
  labs(
    x = "Majority Race",
    y = "Distance to Nearest Stop (miles)",
    title = "Distance to Nearest SEPTA Stop by Racial Composition"
  ) +
  theme_minimal() +
  theme(legend.position = "none")



```


**Your interpretation:**

SEPTA transit stops are unevenly distributed across Philadelphia's census tracts, with notable disparities along racial lines. Majority-Black and No Majority tracts tend to have higher raw stop counts but lower stops per capita, suggesting that while transit infrastructure is physically present, it may not be proportionally serving the population. Majority-White tracts, particularly in the northeast and northwest of the city, show greater average distances to the nearest transit stop, reflecting lower transit density in more car-dependent, suburban-style neighborhoods



---

